E-Graphs With Bindings

### 2025-01-23 LICS_2025/e_graphs_with_bindings/E-graphs-with-bindings.tex
fixed grey color
last section left for closed

\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}

\usepackage{amsthm}

@@

\begin{abstract}
    The technique of equality saturation for program optimization and reasoning has recently gained significant attention due to the renewed popularity of equality graphs (e-graphs).
    E-graphs represent multiple terms that are equivalent modulo a given set of rewrite rules. 
    This capability allows rewriting to be performed over a family of terms simultaneously.    
    However, there are domains where e-graphs fail to achieve optimal performance. Notably, e-graphs struggle in domains involving variable binding, such as $\lambda$-calculus, because they lack native support for binding.
    Building on recent work that provides a categorical interpretation of e-graphs as morphisms in a semilattice-enriched symmetric monoidal category, we extend this framework. Specifically, we show how e-graphs with bindings can be derived as morphisms in a closed symmetric monoidal category with similar enrichment. This extension provides a principled approach to equality saturation for $\lambda$-calculus.
    We then define a concrete combinatorial representation of these morphisms using hierarchical hypergraphs. Additionally, we introduce a corresponding notion of double-pushout (DPO) rewriting. Finally, we demonstrate the equivalence between term rewriting and DPO rewriting, with the key property that DPO rewriting absorbs the equations of the symmetric monoidal category.
\end{abstract}

@@

\section{Introduction}

@@

All these issues can be resolved by the use of \emph{string diagrams}.
Thus, in this paper we take what we consider to be a principled approach to e-graphs with bindings by lifting to monoidal closed categories prior work showing how e-graphs (without binding) can be given a categorical semantics using semilattice-enriched symmetric monoidal categories, which allows e-graphs to be reconstructed as string diagrams.~\cite{ghica2024equivalencehypergraphsegraphsmonoidal}
The approach in \emph{loc. cit.} was originally motivated by the desire to support the machinery of categorial (DPO) graph rewriting for e-graphs, but by extending the semilattice-enrichment to monoidal-categories it paves the way to a natural derivation of e-graphs with binding.

@@

We build on the technology of string diagrams, which is by now well developed in a number of areas. 
String diagrams have the advantage of being a direct bridge from the language of category theory, which in turn can be used to give semantic models to a variety of programming languages and logics, to the concrete language of graphs, all based on solid mathematical foundations. 
To solve the new problems discuss above, of dealing with binding and capture avoiding substitution, we just follow the categorical breadcrumb trail.

@@

E-graphs can efficiently simultaneously represent several equivalent terms of an algebraic theory, generalising the DAG representation of terms to include equivalence classes.
The key concepts of e-graphs can be intuitively grasped from some simple (but non-trivial) examples, as shown in Figure~\ref{fig:e-graph-example}, a `Hello world!' of e-graphs~\cite{EggPaper}.

@@

Before diving into the technical details let us compare how substitution works in conventional e-graphs as compared to our proposed approach.
In Figure~\ref{fig:e-graph-substitution} we show successive iterations of a conventional e-graph for $(\lambda x . (y + y) + x) 1$ following the $\beta$-reductions.
A subterm for an explicit substitution must be introduced, which is deemed equivalent to the top-level application term, to be rewritten using the rules for explicit substitution at the top of each column.
At the end, all explicit substitution nodes can be removed.
The same example is then given again in Figure~\ref{fig:e-string-substitution}, this time using string diagrams.

@@

\input{sections/categorical_semantics}
\input{sections/combinatorial_semantics}

>>>>

\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor,soul}

\usepackage{amsthm}
\usepackage{hyperref}
\usepackage{stmaryrd}

@@

\begin{abstract}
    The technique of equality saturation for program optimisation and reasoning has recently gained significant attention due to the renewed popularity of equality graphs (e-graphs).
    E-graphs represent multiple terms that are equivalent modulo a given set of rewrite rules. 
    This capability allows rewriting to be performed over a family of terms simultaneously.    
    However, there are domains where e-graphs fail to achieve optimal performance. 
	Notably, e-graphs struggle in domains involving variable binding, such as $\lambda$-calculus, because they lack native support for binding.
    Building on recent work that provided a categorical interpretation of e-graphs as morphisms in a semilattice-enriched symmetric monoidal category, we extend this framework. 
	Specifically, we show how e-graphs with bindings can be derived as morphisms in a closed symmetric monoidal category with similar enrichment. 
	This extension provides a principled approach to equality saturation for $\lambda$-calculus.
    We then define a concrete combinatorial representation of these morphisms using hierarchical hypergraphs. 
	Additionally, we introduce a corresponding notion of double-pushout (DPO) rewriting. 
	Finally, we demonstrate the equivalence between term rewriting and DPO rewriting, with the key property that DPO rewriting absorbs the equations of the symmetric monoidal category.
\end{abstract}

@@

\section{Introduction}
\label{sec:introduction}

@@

All these issues can be resolved by the use of \emph{string diagrams}.
Thus, in this paper we take what we consider to be a principled approach to e-graphs with bindings by lifting to monoidal closed categories prior work of~\cite{ghica2024equivalencehypergraphsegraphsmonoidal} showing how e-graphs (without binding) can be given a categorical semantics using semilattice-enriched symmetric monoidal categories, which allows e-graphs to be reconstructed as string diagrams.
The approach in \emph{loc. cit.} was originally motivated by the desire to support the machinery of categorial (DPO) graph rewriting for e-graphs, but by extending the semilattice-enrichment to monoidal-categories it paves the way to a natural derivation of e-graphs with binding.

@@

We build on the technology of string diagrams, which is by now well developed in a number of areas. 
String diagrams have the advantage of being a direct bridge from the language of category theory, which in turn can be used to give semantic models to a variety of programming languages and logics, to the concrete language of graphs, all based on solid mathematical foundations. 
To solve the new problems discusses above, of dealing with binding and capture avoiding substitution, we just follow the categorical breadcrumb trail.

@@

E-graphs can efficiently simultaneously represent several equivalent terms of an algebraic theory, generalising the \textit{directed acyclic graph} (DAG) representation of terms to include equivalence classes of subterms (or subgraphs).
The key concepts of e-graphs can be intuitively grasped from some simple (but non-trivial) examples, as shown in Figure~\ref{fig:e-graph-example}, where it is shown how a series of rewrite rules can be non-destructively applied to an e-graph as first presented in ~\cite{EggPaper}.

@@

Before diving into the technical details let us compare how substitution works in conventional e-graphs as compared to our proposed approach.
In Figure~\ref{fig:e-graph-substitution} we show successive iterations of a conventional e-graph for $(\lambda x . (y + y) + x) 1$ following the $\beta$-reductions.
A subterm for an explicit substitution must be introduced, which is deemed equivalent to the top-level application term (hence they are inside the same dashed box), to be rewritten using the rules for explicit substitution at the top of each column.
At the end, all explicit substitution nodes can be removed.
The same example is then given again in Figure~\ref{fig:e-string-substitution}, this time using string diagrams.

@@

\input{sections/categorical_semantics}
\input{sections/combinatorial_semantics}
\input{sections/soundness}

### 2025-01-23 LICS_2025/e_graphs_with_bindings/sections/categorical_semantics.tex
last section left for closed

We recall some basic definitions.
\begin{definition}
A \textit{monoidal} \textit{signature} $\Sigma = (\Sigma_{O}, \Sigma_{M}, t)$ is given by a set of objects (called types) $\Sigma_{O}$, a set of generators (operations) $\Sigma_{M}$ and a type assignment function $t : \Sigma_{M} \to \Sigma_{O}^{*} \times \Sigma_{O}^{*}$ assigning lists of input and output types to a given operation.
\end{definition}

@@

\begin{definition}
The free closed symmetric monoidal category $\catname{CS}(\Sigma)$ is given by a set of closed $\Sigma$-terms quotiented by the laws of a symmetric monoidal category and the equations from Definition~\ref{def:closed}.
\end{definition}

To encode equivalences between morphisms (or $\Sigma$-terms), the hom-sets of the corresponding free categories can be endowed with the additional structure of a semilattice induced by an idempotent, commutative and associative operator $+$.
This endowment turns such categories into $\catname{SLat}$-enriched categories where $\catname{SLat}$ is a category of \textit{unbounded} semilattices and homomorphisms between them (see Appendix section~\ref{sec:appendix:slat}).

@@

In particular, 2-functor $\mathcal{F}$ turns every category $\mathbb{C'} \in \catname{Cat}$ into a free semilattice enriched $\mathbb{C} \in \catname{SLat}\text{--}\catname{Cat}$ by making every hom-set of $\mathbb{C'}$ into a free semilattice on this set.
Being a 2-functor it also lifts an adjunction in $\catname{Cat}$ $F \dashv G$ into an $\catname{Slat}$-adjunction.
Thus, applying the functor to $\catname{CS}(\Sigma)$ we get a closed symmetric monoidal $\catname{Slat}$-category $\catname{CS}(\Sigma)^{+}$ by lifting the adjunction $(- \otimes B, B \multimap{} -, \eta, \varepsilon)$.
We can similarly derive $\catname{SLat}$-enriched versions of  $\Lambda_{A,B,C}$ and $\textbf{ev}_{A,B}$ morphisms using the lifted adjunction as $\textbf{ev}_{A,B} = \varepsilon_{A}$ and $\Lambda_{A,B,C}(f) = \eta; (B \multimap{} f)$.
The following distributivity law can then be derived, forming the basis of encoding equivalences inside abstractions.
\begin{align*}
\Lambda_{A,B,C}(f + g) &= \eta;(B \multimap (f + g))\\
                       &= \eta;(B \multimap (f) + B \multimap (g))\\
                       &= \eta;(B \multimap (f)) + \eta;(B \multimap (g))\\
                       &= \Lambda_{A,B,C}(f) + \Lambda_{A,B,C}(g)
\end{align*}

Similar to $\catname{CS}(\Sigma)$, $\catname{CS}(\Sigma)^{+}$ can be freely generated from a set of $\Sigma^{+}$-terms, the latter being defined as follows.
Every $\Sigma$-term is also a $\Sigma^{+}$-term. 
Given two $\Sigma^{+}$-terms $f : A \to B$ and $g : A \to B$, $f + g : A \to B$ is also a $\Sigma^{+}$-term.

To aid reasoning, we introduce a new language of string diagrams for $\catname{SLat}$-SMCs, using a hierarchical ``box'' structure to capture the join operation on morphisms.  It was already discussed in the introduction how this is  used in the translation of equivalence classes from the e-graph to the string diagrammatic setting. 
Figure \ref{fig:egraph-strings} displays the generators of this language which is the usual string diagrammatic syntax~\cite{Selinger_2010} apart from the last component which is our notation for $+$; the first component denotes an empty diagram. 
Figure \ref{fig:string-equations} displays the additional equations which these diagrams satisfy, in addition to the standard SMC equations. 
The first four equations are those displayed in Definition \ref{def:enriched-prop},  while the final four axiomatize $+$ as an \textit{n-ary} associative, commutative and idempotent operation.  We overload the binary notation $+$ for our $n$-ary notation.  
We will later prove the intuitive fact that these diagrams are sound and complete with respect to their intended categorical semantics, noting that similar diagrammatic languages using boxes to express choice have been used before~\cite{duncan_generalised_2009}.

>>>>

Given a category $\mathbb{C}$  with objects $A,B \in \mathbb{C}$ we denote by $\mathbb{C}(A,B)$ the corresponding hom-set.  
We write the identity morphism on $A$ as $\id_A$.  
We commonly write $f;g$ for composition in diagrammatic order.  
Composition in the usual order is written $g \circ f$.  
We denote the tensor product of an SMC $\mathbb{C}$ by $\otimes$,  its unit by $I$, and its symmetry natural transformation as $\sym$ \cite{maclane}.  
We adopt the convention that $\otimes$ binds more tightly than $(;\!)$.  
We elide all associativity and unit isomorphisms associated with monoidal categories,  and often omit subscripts on identities and natural transformations where it can be inferred.

First, recall some basic definitions.
\begin{definition}
A \textit{monoidal} \textit{signature} $\Sigma = (\Sigma_{O}, \Sigma_{M}, t)$ is given by a set of objects (called types) $\Sigma_{O}$, a set of generators (operations) $\Sigma_{M}$ and a type assignment function $t : \Sigma_{M} \to \Sigma_{O}^{*} \times \Sigma_{O}^{*}$ assigning lists of input and output types to a given operation.
\end{definition}

@@

\begin{definition}
The free closed symmetric monoidal category $\catname{CS}(\Sigma)$ is given by a set of closed $\Sigma$-terms quotiented by the laws of a symmetric monoidal category and the equations from Definition~\ref{def:closed}.
We can similarly consider a free closed symmetric monoidal category that is induced by an $\catname{SMT}(\Sigma, \mathcal{E})$ over closed $\Sigma$-terms which we will denote with $\catname{CS}(\Sigma, \mathcal{E})$.
\end{definition}

To encode equivalences between morphisms (or $\Sigma$-terms), the hom-sets of the corresponding free categories can be endowed with the additional structure of a semilattice induced by an idempotent, commutative and associative operator \textit{join} $+$ as presented in~\cite{ghica2024equivalencehypergraphsegraphsmonoidal}.
This endowment turns such categories into $\catname{SLat}$-enriched categories where $\catname{SLat}$ is a category of \textit{unbounded} semilattices and homomorphisms between them (see Appendix section~\ref{sec:appendix:slat}).

@@

In particular, 2-functor $\mathcal{F}$ turns every category $\mathbb{C'} \in \catname{Cat}$ into a free semilattice enriched $\mathbb{C} \in \catname{SLat}\text{--}\catname{Cat}$ by making every hom-set of $\mathbb{C'}$ into a free semilattice on this set.
Being a 2-functor it also lifts an adjunction in $\catname{Cat}$ $F \dashv G$ into an $\catname{Slat}$-adjunction.
Thus, applying the functor to $\catname{CS}(\Sigma)$ we get a closed symmetric monoidal $\catname{Slat}$-category $\catname{CS}(\Sigma)^{+}$ by lifting the adjunction $(- \otimes B, B \multimap{} -, \eta, \varepsilon)$.
We can similarly derive $\catname{SLat}$-enriched versions of  $\Lambda_{A,B,C}$ and $\textbf{ev}_{A,B}$ morphisms using the lifted adjunction as $\textbf{ev}_{A,B} = \varepsilon_{A}$ and $\Lambda_{A,B,C}(f) = \eta; (B \multimap{} f)$.
The following distributivity law can then be derived, forming the basis of encoding equivalences inside abstractions.
\begin{equation}
\begin{aligned}
\Lambda_{A,B,C}(f + g) &= \eta;(B \multimap (f + g))\\
                       &= \eta;(B \multimap (f) + B \multimap (g))\\
                       &= \eta;(B \multimap (f)) + \eta;(B \multimap (g))\\
                       &= \Lambda_{A,B,C}(f) + \Lambda_{A,B,C}(g)
\end{aligned}
\end{equation}
\label{law:distributivity}

Similar to $\catname{CS}(\Sigma)$, $\catname{CS}(\Sigma)^{+}$ can be freely generated from a set of $\Sigma^{+}$-terms, the latter being defined as follows.
Every $\Sigma$-term is also a $\Sigma^{+}$-term. 
Given two $\Sigma^{+}$-terms $f : A \to B$ and $g : A \to B$, $f + g : A \to B$ is also a $\Sigma^{+}$-term.
Terms of $\catname{CS}(\Sigma)^{+}$ are quotiented by the laws of closed $\catname{SLat}$-SMC, including the distributivity law~\ref{law:distributivity}.

To aid reasoning, we introduce a new language of string diagrams for closed $\catname{SLat}$-SMCs, using a hierarchical ``box'' structure to capture the join operation on morphisms as well as another hierarchical ``box'' for lambda abstraction.
It was already discussed in the introduction how this is  used in the translation of equivalence classes from the e-graph to the string diagrammatic setting and in interpreting $\lambda$-terms.
Figure \ref{fig:egraph-strings} displays the generators of this language which is the usual string diagrammatic syntax~\cite{Selinger_2010} plus the syntax for lambda abstraction as established in~\cite{ghica2024stringdiagramslambdacalculifunctional} and syntax for semilattice-enrichment as introduced in~\cite{ghica2024equivalencehypergraphsegraphsmonoidal}; the first component denotes an empty diagram.
Note the orientation of string diagrams are now from top to bottom as opposed to what was used in~\nameref{sec:introduction}: the latter was mimicking the orientation of e-graphs, \textit{i.e.}, traditional orientation for expression trees.
% Figure \ref{fig:string-equations} displays the additional equations which these diagrams satisfy, in addition to the standard SMC equations. 
% The first four equations are those displayed in Definition \ref{def:enriched-prop},  while the final four axiomatize $+$ as an \textit{n-ary} associative, commutative and idempotent operation.  We overload the binary notation $+$ for our $n$-ary notation.  
% We will later prove the intuitive fact that these diagrams are sound and complete with respect to their intended categorical semantics, noting that similar diagrammatic languages using boxes to express choice have been used before~\cite{duncan_generalised_2009}.

In~\cite{ghica2024stringdiagramslambdacalculifunctional} it was shown that e-graphs over some signature $\Sigma$ can be represented as morphisms of a semilattice-enriched free Cartesian symmetric monoidal category over the same signature.
We argue that restricting the categorical domain to free semilattice-enriched closed symmetric monoidal categories gives rise to structures that naturally support binding and equivalence classes of morphisms.
That is, morphisms (string diagrams) of such category can both encode variable binding and equivalence between subterms (sub-diagrams).
We call these string diagrams, or more precisely, the combinatorial representation of that will follow, e-graphs with bindings.

\begin{figure*}
\[
\adjustbox{scale=0.75}{
\tikzfig{./figures/egraph-strings}
}
\]
\caption{String diagrams for closed semilattice-enriched symmetric monoidal categories.}
\label{fig:egraph-strings}
\end{figure*}
